The Way Forward in Parallelizing Dynamic Languages

### 2014-04-29 talk/icooolps2014/position-paper.tex
more text

\section{Discussion}
(why do we have the GIL? it makes the interpreter thread-safe)
(now there is code relying on the semantics (atomicity of bytecode instructions))
(synchronization with locks for application-level synchronization is still needed. So GIL is useless for that)

>>>>

\section{Discussion}
\subsection{Why is there a GIL?}
The GIL is a very simple synchronization mechanism for supporting
multi-threading in the interpreter. The basic guarantee is that the
GIL may only be released in-between bytecode instructions. The
interpreter can thus rely on complete isolation and atomicity of these
instructions. As a consequence, applications can rely on certain
operations to be atomic. While this is probably not a good idea,
it is used in practice. A solution replacing the GIL should therefore
uphold these guarantees, while preferably also be as easily
implementable as a GIL for the interpreter.

The GIL also allows for easy integration with external C libraries that
do not need to be thread-safe. For the duration of the calls, we
simply do not release the GIL. External libraries that are explicitly
thread-safe can voluntarily release the GIL themselves in order to
still provide some parallelism. This is done for example for
potentially long I/O operations. Consequently, I/O-bound,
multi-threaded applications can actually parallelize to some
degree. Again, a potential solution should be able to integrate with
external libraries with similar ease. We will however focus our
argumentation more on running code in the interpreted language in
parallel, not the external C calls.

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronize memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is well
known that using locks for synchronization is not easy.  They are
non-composable, have overhead, may deadlock, limit scalability, and
overall add a lot of complexity. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronization mechanism called \emph{atomic blocks}.

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions.  Interpreters using using a GIL can simply
guarantee that the GIL is not released during the execution of the
atomic block. Of course, this still means that no two atomic blocks
can execute in parallel or even concurrently. Potential solutions
that provide a good way to implement atomic blocks are therefore
preferable.

\subsection{Potential Solutions}
